---
title: "Empirical Finance: Tutorial - Week 1"
author: "Tomás Cordeiro da Silva"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    self_contained: true
---

# Setup 

## Working Directory
```{r, warning=FALSE, message=FALSE}
# --- confirm working directory
getwd()

# --- list existing files
list.files()
```

## Housekeeping
```{r, warning=FALSE, message=FALSE}
# --- removes all objects from the the current environment
rm(list = ls())

# --- removes a specific object 
# rm(my_object)

# --- removes a specific object stored as a string
# rm(list = "my_object")
```

# Functions
For most tasks R has a function for it. When it does not, most likely someone else already created one and made it publicly available via a package. Still, you can come across a situation where you will need to build your own. Refer to the "R Primer" document to learn how to create your own.

# Reading Data

## API
An API is a tool that allows you to easily and automatically access data without having to download it. Particularly useful when dealing with large datasets or ones that update regularly.

An example of an API is the `wbstats` package, which provides direct access to World Bank Indicators.
```{r, warning=FALSE, message=FALSE}
library(wbstats)

gdp_data <- wb_data(
  country = "GBR",
  indicator = "NY.GDP.PCAP.CD"
  )

# --- we won't be needing this data today, so we can just remove it 
rm("gdp_data")
```

## `read.csv` 
On the "R Primer" document, I used `read.table`, instead of `read.csv`. `read.table` also reads csv files, as long as you specify that the separator as a comma (,). `read.csv` takes the comma as the default separator.

Let's now load the data we will work with in this session.
```{r, warning=FALSE, message=FALSE}
# --- loading the data 
data <- read.csv("Data/eaef21.csv")

# --- how does the data look like?
head(data)
```

Some initial data assessment.
```{r, warning=FALSE, message=FALSE}
# --- what are the dimensions of our dataset?
dim(data)
# -- number of rows
nrow(data)
# -- number of columns
ncol(data)

# --- how much memory does this object occupy in R?
object.size(data)
```

```{r, warning=FALSE, message=FALSE}
# --- what are the variables in our dataset?
colnames(data)
```

```{r}
# --- what is the class of our variables (i.e. how is the object treated)?
sapply(data, class)
# -- EARNINGS will latter be our variable of interest
class(data$EARNINGS)
```

```{r}
# --- getting the pulse of the variables 
summary(data)
# -- again, let's us focus on EARNINGS
summary(data$EARNINGS)
```

# R Packages 

## How to install a package 
```{r, warning=FALSE, message=FALSE}
# --- command to install a package
# install.packages("tidyverse")

# --- once the package is installed 
library(tidyverse)

# --- listing the packages that come tidyverse
tidyverse::tidyverse_packages()
```

## `dplyr` package 
```{r, warning=FALSE, message=FALSE}
# --- create new variables with mutate
data <- data %>%
  mutate(
    EDUC_LEVEL= factor(
      case_when(
        EDUCDO == 1 ~ "low",
        EDUCHSD == 1 | EDUCPROF == 1 ~ "medium",
        TRUE ~ "high"
      ),
      levels = c("low", "medium", "high")
    )
  )

# --- perform operations by cluster with group_by and summarise
data %>%
  group_by(MALE) %>%
  summarise(mean(EARNINGS), mean(HOURS), mean(EXP))

# --- perform multiple operations at the same time 
data %>% 
  mutate(AGE_40 = ifelse(AGE <= 40, "below 40 or 40", "above 40")) %>%
  group_by(AGE_40, MALE) %>% 
  summarise(mean(EARNINGS), mean(HOURS), mean(EXP))
```

## `ggplot` package
```{r, warning=FALSE, message=FALSE}
ggplot(data = data, aes(x = EXP, y = EARNINGS)) + 
  geom_point()
```

We can spice up our plot with these add-ons.
```{r, warning=FALSE, message=FALSE}
ggplot(data = data, aes(x = EXP, y = EARNINGS)) +
  # --- mapping the color of the points to the education level, generates  distinct colors (which will be chosen by R)
  geom_point(aes(color = EDUC_LEVEL)) + 
  # --- we can modify the colors and labels to our choosing
  scale_color_manual(
    values = c(
      "low" = "#C44E52", 
      "medium" = "steelblue",
      "high" = "#4E9A06"
      ),
    name = "education level"
  ) +
  # --- we can write the labels of our axis and the title of our plot
  labs(
    x = "EXPERIENCE",
    y = "EARNINGS",
    title = "scatter plot") +
  # --- to remove the grey background colour we just call this option
  theme_minimal() + 
  # --- for further details 
  theme(
    # -- options for axis text and title
    axis.text.x  = element_text(color = "black"),
    axis.text.y  = element_text(color = "black"),
    axis.title.x = element_text(color = "black"),
    axis.title.y = element_text(color = "black"),
    # -- option to mark a line in both axis
    axis.line = element_line(color = "black"),
    # -- option to include ticks in both axis
    axis.ticks.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black")
  )
```

Another informative plot.
```{r, warning=FALSE, message=FALSE}
ggplot(data = data, aes(x = EARNINGS, fill = EDUC_LEVEL)) + 
  geom_density(
    # -- alpha controls the transparency of the area under the curve
    alpha = 0.4,
    color = "black"
    ) + 
  scale_fill_manual(
    values = c(
      "low" = "#C44E52", 
      "medium" = "steelblue",
      "high" = "#4E9A06"
      ),
    name = "education level"
  ) +
  scale_y_continuous(
    expand = c(0,0)
  ) +
  labs(
    x = "EARNINGS",
    y = "DENSITY",
    title = "density of EARNINGS by EDUCATION LEVEL"
  ) +
  theme_minimal() +
  # --- for further details 
  theme(
    # -- options for axis text and title
    axis.text.x  = element_text(color = "black"),
    axis.text.y  = element_text(color = "black"),
    axis.title.x = element_text(color = "black"),
    axis.title.y = element_text(color = "black"),
    # -- option to mark a line in both axis
    axis.line = element_line(color = "black"),
    # -- option to include ticks in both axis,
    axis.ticks.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
    # -- remove vertical grid lines
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```

# Regressions 

## Univariate Regression
```{r}
model_simple <- data %>%
  lm(log(EARNINGS) ~ S, data =.)

model_simple %>% summary()
```

## Multivariate Regression
```{r}
model_multi <- data %>%
  lm(log(EARNINGS) ~ S + EXP + MALE, data = .)

model_multi %>% summary()
```

```{r}
model_educ_group <- data %>%
  lm(log(EARNINGS) ~ EDUC_LEVEL + EXP + MALE, data = .)

model_educ_group %>% summary()
```
